---
title: "Fiji: An Overview"
author: "Jin Wen Lin"
format: pdf
editor: visual
---

```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(readxl)
```


## Introduction  

Fiji is a tropical island country located in Oceania, the South Pacific Ocean. It is located about two-thirds of the way between Hawaii and New Zealand. In addition, it is made up of over 332 islands, with 110 permanently inhabited. The three islands that are close to Fiji are Vanuatu, Samoa, and Tonga.  

## Key Facts  

### Government  

The government of Fiji is a republic with a president as the Chief of State and a prime minister as the Head of the Government. The current president of Fiji is Ratu Wiliame Katonivere, and the prime minister is Sitiveni Ligamamada Rabuka. Its political system is based on a parliamentary democracy, where members of parliament are elected in order to make decisions on behalf of the people. President is elected by the parliament indirectly and the prime minister is elected by the parliament. The election cycle is 3 years. The highest level of judicial authority in Fiji is the federal supreme court where judges work independently and follow the law in order to make decisions.  

### Economy  

Fiji is described as one of the most developed countries in the Pacific region. The top industries of Fiji are tourism, sugar, clothing, and copra (where coconut oil is extracted). Fiji's largest trading partners are Australia, United States, and New Zealand, and the exports include beverages, oil and mineral fuels, and seafood. The largest composition of GDP (Gross Domestic Product) is in services, followed by industry, manufacturing and agriculture.  

The following graph illustrates Fiji's GDP over the years. This shows how the economic 
output has changed over time.  

```{r, echo=FALSE}
# gdp data
gdp_data <- read_excel("imf-dm-export-20241213.xls")
gdp_clean <- gdp_data %>% 
  filter(`GDP, current prices (Billions of U.S. dollars)`== "Fiji") %>% 
  pivot_longer(
    cols = -`GDP, current prices (Billions of U.S. dollars)`,
    names_to = "Year",
    values_to = "GDP"
  ) %>%
  select(Year, GDP) %>%
  mutate(Year = as.numeric(Year), GDP = as.numeric(GDP)) %>% 
  filter(Year < 2024) # years after 2024 are gdp projections
```


```{r, echo=FALSE}
# graph of gdp
ggplot(gdp_clean, aes(x = Year, y = GDP)) + 
      geom_line(color = "darkblue") + 
      geom_point(color = "blue") + 
      labs(title = "GDP of Fiji Over Years",
           x = "Year", y = "GDP (Billions of U.S. dollars)") + 
      theme_minimal() 
```

The graph above shows an increasing trend in Fiji's GDP over time. Before the year of 2000, 
the GDP increases gradually but the GDP starts to rise more sharp after 2000, which indicates 
a steady economic growth. Additionally, this graph also shows some fluctuations, which may 
suggests some economic events or challenges that impacted the economic growth. For 
example, there is a sharp decrease in GDP around the years of 2019 to 2020, which might 
be impacted by the issue of Covid-19.  

### Population  

The population in 2024 is approximately 928,784, with the median age of 27.9, which is quite young. Of the population, about 60% of the people live in the urban area of Fiji.  

The following graph shows the total population of Fiji over years.  

```{r, echo=FALSE}
# population data
population_data <- read.csv("POPTOTFJA647NWDB.csv")
# date format
population_data$observation_date <- as.Date(population_data$observation_date, 
                                            format = "%m/%d/%Y")
population_data$Year <- year(population_data$observation_date)
population_data <- population_data %>% 
  rename(Population = POPTOTFJA647NWDB)
```


```{r, echo=FALSE}
# graph of population
ggplot(population_data, aes(x = Year, y = Population)) + 
      geom_line(color = "purple") + 
      geom_point(color = "violet") + 
      labs(title = "Population of Fiji Over Years",
           x = "Year", y = "Population") + 
      theme_minimal() 
```

Overall, the above graph shows an steady increasing trend of the total population in Fiji. This tend may suggests that the living standards and healthcare etc. are improved over the years. It is easily to see that the population in 2020 is over 800,000, where the population in 1960 is only around 400,000. The population is approximately doubled over the years, hence the development of Fiji is getting better.  

### Tourism  
Fiji is a popular tourist destination known for its beautiful beaches, tropical climate, 
and clean waters. People are enjoying doing activities like exploring the natural environment and diving etc. at Fiji.  

The following graph shows the total number of visitors to Fiji over years.

```{r, echo=FALSE}
# visitor data
visitor_data <- read.csv("summary-of-visitor-arriv.csv")
visitor_data <- visitor_data %>% 
  rename(Year = Category) 
# From another research showing the visitors in 2023
visitor_data <- add_row(visitor_data, Year = 2023, Arrivals = 929740)
visitor_data <- subset(visitor_data, select = -Departures)
```


```{r, echo=FALSE}
# graph of visitors
ggplot(data = visitor_data, aes(x = Year, y = Arrivals)) + 
      geom_line(color = "darkgreen") + 
      geom_point() + 
      labs(title = "Visitors to Fiji", y = "Number of Visitors", x = "Year") + 
      theme_minimal()
```

There is an increasing pattern in the number of visitors to Fiji start from 2000 to 2019. 
However, there is a sharp decrease between the years of 2019 to 2021. This is most likely due 
to the reason of Covid-19. After 2021, the number of visitors increases very quickly 
hence suggesting that the tourism is recovering.  

## Projection  

### Population  

A simple linear model with response variable population and time as the predictor is fitted for the 10 years projection. Here is the linear model.  

```{r, echo=FALSE}
population_model <- lm(Population ~ Year, data = population_data)
summary(population_model)
```
Now, here is the projection based on this model. 

```{r, echo=FALSE}
# projection data
future_years <- (max(population_data$Year) + 1):(max(population_data$Year) + 10)
future_population <- predict(population_model, newdata = data.frame(Year = future_years))
projection_data <- data.frame(Year = future_years, 
                       Projected = future_population)

# projection plot
ggplot(projection_data, aes(x = Year, y = Projected)) +
      geom_line() +
      geom_point() +
      labs(title = "Population Projection",
           x = "Year", y = "Projected Value") +
      theme_minimal()
```

### GDP  

Here is the simple linear model for GDP and projection is made using this fitted model.  

```{r, echo=FALSE}
gdp_model <- lm(GDP ~ Year, data = gdp_clean)
summary(gdp_model)
```

Here is the GDP projection for the next 10 years.  

```{r, echo=FALSE}
# projection data
future_years <- (max(population_data$Year) + 1):(max(population_data$Year) + 10)
future_gdp <- predict(gdp_model, newdata = data.frame(Year = future_years))
projection_data <- data.frame(Year = future_years, 
                       Projected = future_gdp)

# projection plot
ggplot(projection_data, aes(x = Year, y = Projected)) +
      geom_line() +
      geom_point() +
      labs(title = "GDP Projection",
           x = "Year", y = "Projected Value") +
      theme_minimal()
```

### Visitors  

The following is the fitted model for the projection of visitors.  

```{r, echo=FALSE}
visitors_model <- lm(Arrivals ~ Year, data = visitor_data)
summary(visitors_model)
```

Here is the projection of visitors in the next 10 years.  

```{r, echo=FALSE}
# projection data
future_years <- (max(population_data$Year) + 1):(max(population_data$Year) + 10)
future_visitors <- predict(visitors_model, newdata = data.frame(Year = future_years))
projection_data <- data.frame(Year = future_years, 
                       Projected = future_visitors)

# projection plot
ggplot(projection_data, aes(x = Year, y = Projected)) +
      geom_line() +
      geom_point() +
      labs(title = "Visitors Projection",
           x = "Year", y = "Projected Value") +
      theme_minimal()
```

## Comparison  

The islands that Fiji are comparing to are Vanautu, Samoa, and Tonga. As mentioned, they are the three neighbors of Fiji.  

### Population  

Here is the comparison of population in the year of 2024 between the islands.  

```{r, echo=FALSE}
# comparison data
comparison_data <- data.frame(
  Island = c("Fiji", "Vanuatu", "Samoa", "Tonga"),
  Population = c(936375, 327777, 218019, 104175),
  GDP = c(5.442, 1.13, 0.93, 0.52),  # in billion USD
  Visitors = c(929740, 30306, 241276, 119916))

# comparison plot
ggplot(comparison_data, aes(x = reorder(Island, -Population), 
                            y = Population, fill = Island)) +
      geom_bar(stat = "identity", color = "black") +
      labs(title = "Comparison of Population among Island States",
           x = "Island State", y = "Population") +
      scale_fill_brewer(palette = "Set2") +
      theme_minimal()
```

Population: By comparing the population of Fiji with its neighbour islands Vanautu, Samoa, and Tonga, Fiji has the largest population. Vanuatu ranks the second and followed closely by Samoa. They have much smaller population compared to Fiji. Lastly, Tonga has the smallest population.  

### GDP  

Here is the comparison for GDP.  

```{r, echo=FALSE}
# comparison plot
ggplot(comparison_data, aes(x = reorder(Island, -GDP), 
                            y = GDP, fill = Island)) +
      geom_bar(stat = "identity", color = "black") +
      labs(title = "Comparison of GDP among Island States",
           x = "Island State", y = "GDP") +
      scale_fill_brewer(palette = "Set2") +
      theme_minimal()
```

By comparing the 2023 GDP of Fiji with its neighbours, Fiji dominates the Pacific region 
economically. This may indicates that Fiji has a more developed economy compared to its neighbours. Then, followed by Vanuatu and Samoa, with Vanuatu has slightly higher GDP. Lastly, Tonga has the lowest GDP among the islands shown.  

### Visitors  

Here is the comparison for the visitors.  

```{r, echo=FALSE}
# comparison plot
ggplot(comparison_data, aes(x = reorder(Island, -Visitors), 
                            y = Visitors, fill = Island)) +
      geom_bar(stat = "identity", color = "black") +
      labs(title = "Comparison of Visitors among Island States",
           x = "Island State", y = "Visitors") +
      scale_fill_brewer(palette = "Set2") +
      theme_minimal()
```

By comparing the 2023 visitors of Fiji with its neighbour islands, again Fiji has a significant lead in the number of visitors. Samoa ranked the second, and is far behind Fiji. 
Then, it is followed by Tonga ranked third place and the last place Vanuatu. This may suggests that the tourism sector for Vanuatu is less popular compared to other islands.  

## References  

**Data Sources:**  

https://fred.stlouisfed.org/series/POPTOTFJA647NWDB  

https://www.imf.org/external/datamapper/NGDPD@WEO/FJI?zoom=FJI&highlight=FJI  

https://www.statsfiji.gov.fj/statistics/social-statistics/tourism-and-migration-statistics/  

**Research References:**  

https://www.investinfiji.today/news-publication/fiji-breaks-visitor-arrival-records/#:~:text=Fiji%20welcomed%20a%20record%20929%2C740,Canada%20have%20exceeded%20past%20years.  

https://www.worldometers.info"  

https://tradingeconomics.com/  

https://globaledge.msu.edu/countries/fiji  

https://www.sbs.gov.ws/documents/social/International_Arrival_Statistics/2023/Migration_Annual_23.pdf  

https://www.statista.com/statistics/729470/gross-domestic-product-gdp-in-tonga/  

https://tongastats.gov.to/statistics/social-statistics/migration/  



